Thermal reactionomes reveal adaptive responses to thermal extremes in warm

### 2014-10-23 doc/ApTranscriptome_manuscript.Rmd

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we describe a novel analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis* in a heat-shock experiment. We found that at least 8% of all genes were thermally responsive. The majority of upregulation was specific to low temperatures, and consistent with the *enhanced defense hypothesis* the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptation. To better understand the relationship between these two mechanisms, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. We found substantial differences in their patterns of gene expression. *Aphaenogaster picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors.

@@

We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced defense hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced responses to provide greater protection from more frequently encountered stressors. This hypothesis would predict that the cool-adapted *A. picea* would induce more genes in response to low temperatures than would the warm-adapted *A. carolinensis*, which would induce more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should also be stronger for more commonly experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more frequent stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, whereas transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Finally, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014]. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation as the proportion of the transcriptome matching each of these predicted patterns.

@@

pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
             big.mark = ",", split.tables = Inf, 
             caption = "Table of the number of thermally-responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species.")
```

@@

In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species. Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another. These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Table 1, Figure S1) indicating that these differences are not simply due to neutral processes. **Intermediate** transcripts in *A. carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A. picea* than expected by chance (Fig. 1A), whereas there was a significant deficit of the reverse shift of *A. picea* **Intermediate** transcripts to any upregulated category in *A. carolinensis* (Fig. 1B). Similarly, both **High** and **Bimodal** transcripts in *A. picea* were more likely to be **Not Responsive** in *A. carolinensis* than expected, while the reverse was not significant (Fig. 1).

![Deviations from expected numbers of transcripts in matched observation data for transcript expression type data. The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit). Shown are **Low**, **High** and **Bimodal** expression types for (A) *A. picea** and (B) *A. carolinensis* against all expression types in the opposite species. The dashed boxes indicate the diagonals (e.g. High - High) which are positive](../results/matched_observations_Fig1.png)

The molecular processes in the set of transcripts that switched from down-regulation in *A. carolinensis* to up-regulation in *A. picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism. In addition, there were GO terms related to repair (e.g. histone modification), transcript regulation (e.g. positive regulation of cellular process) and cellular organization (e.g. transport, organelle organization). Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A. carolinensis*, primarily shuts down metabolism whereas the cooler-climate species, *A. picea*, increases metabolism and initiates damage repair responses.

We tested the *genetic assimilation hypothesis*, a shift from inducible to constitutive expression of stress-response genes, by examining whether the suite of transcripts induced in response to either temperature extreme showed higher constitutive expression, here defined as expression at 19.5°C, in the species better adapted to these conditions. This would be evident as a positive relationship between the ratio of inducibility between species against the opposite ratio of constitutive expression between species. However, for both high temperature activated transcripts in *A. picea* (Fig. 2A) and low temperature activated transcripts in *A. carolinensis* (Fig. 2B) there was no evidence for greater constitutive expression for transcripts with lower inducibility. In fact, for both comparisons there was the opposite relationship as predicted by the *genetic assimilation hypothesis*, with greater inducibility for transcripts with lower constitutive expression in the opposite species.

![Scatterplots of (A) log ratio of inducibility in *A. picea* over inducibility in *A. carolinensis* against log ration of constitutive expression in *A. carolinensis* over constitutive expression in *A. picae* for transcripts with **High** expression in *A. picea*, and (B) log ratio of inducibility in *A. carolinensis* over inducibility in *A. picea* against log ration of constitutive expression in *A. picea* over constitutive expression in *A. carolinensis*. In both cases, a positive relationship would be Lines are ordinary least squares regressions on the data.](../results/genetic_assimilation_Fig2.png)

@@

![Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures. Red bars and lines are for *A. carolinensis* while blue bars and lines are for *A. picea*.](../results/crit_temp_regulation_Fig3.png)

@@

@@

The potential for many species to persist in face of climate change will depend in part upon their thermal tolerances. However, for most species little is known about the plasticity or adaptive changes in gene expression that underlie thermal tolerances. By using the novel *reactionome* approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient. In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A. carolinensis* we found patterns consistent with different mechanisms of thermal tolerance in these two species.

Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperature extremes: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig. 2A). Moreover, the transcripts upregulated in *A. picea* included nearly the entire set upregulated in *A. carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* might need to induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the short growing season that would necessitate foraging under a broader range of temperatures.

In contrast to cold tolerance, the enhanced upper thermal limit in *A. picea* is best explained by the *tolerance* hypothesis. High temperatures were associated with significantly fewer upregulated transcripts in *A. carolinensis*, and a large proportion (361 of 1,238) of those transcripts upregulated in *A. picea* were either downregulated or expressed at negligible levels overall in *A. carolinensis*. Similarly, signficantly higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* towards the high thermal limit (Fig. 2B, D), suggesting that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases.

The one hypothesis that did not receive unequivocal support was the *genetic assimilation hypothesis*, which predicted that exposure to more frequent stressors is hypothesized to select for a shift from inducible to constitutive expression of stress-response transcripts. In contrast to this prediction, constitutive expression levels at 19.5°C were higher in *A. carolinensis* than *A. picea* in nearly all comparisons, with the specific notable exception of **High** transcripts in *A. picea* that were **Not Responsive** in *A. carolinensis* not because their expression was enhanced, but because they were not expressed at all. This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013], although in a short-term selection experiment for heat tolerance, Sikkink et al. [-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations. Both the genetic assimilation and enhanced defense routes routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair. Whether a particular taxon evolves one strategy over another may be related to the intensity, frequency and duration of temperature stress in a given environment.

Collectively, our results indicate that these two species have distinct mechanisms of response to thermal extremes. The cool-adapated species *Aphaenogaster picea* mounts a classical cellular stress response [@kultz2005; @richter2010] with elevated gene expression at both low and high temperatures, which has been shown to impose a metabolic cost [@hoekstra2013]. In contrast, the warm-adapted *A. carolinensis* relies upon other constitutive defenses [@barshis2013], and perhaps enhanced proteome stability [@ghosh2010] or behavioral responses [@sunday2014] to tolerate thermal stress. Although the reactionome approach is designed to identify large-scale response patterns, this difference between species is also evident among candidate genes known to be involved in stress response. Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A. picea* whereas only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*. These differences are in line with the expectations if *A. picea*, a northern species with fewer good foraging days, must forage in suboptimal conditions whereas *A. carolinensis* can, with a longer season, afford to restrain from foraging in suboptimal conditions. In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

Given the differences in the patterns of thermal responsiveness between species (Fig. 2), it is worth noting the similarity in expression patterns between species. In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. Among closely-related *Drosophila* species, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. To the extent that the number of thermally responsive genes is related to additive genetic variation, our results also suggest that adaptability to high temperatures is more genetically limited than low temperatures, perhaps placing these northern species at risk of competitive replacement by more heat-tolerant southern species [@rodriguez-cabal2012]. At the same time, critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.  Only 15-20% of transcripts upregulated in response to temperature are bimodal (i.e. generalized stress response), and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig. 2). This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.

@@

One possible explanation for the differences between species in their thermal reactionomes is neutral drift of gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009]. While we cannot rule this possibility out, two lines of evidence make us think that it is unlikely. First, if gene expression differed by drift between species, we would not expect to see any striking differences among expression type classes as we do here (Fig. 1, Fig. S1). Second, even if these differences were due to drift, it would be unlikely that there would be transcriptome-wide differences in the critical temperatures of gene activation such as we see here (Fig. 2).

@@

The raw Illumina reads are available from the NCBI Short Read Archive .... The script for analysis and version history is available at (http://github.com/johnstantongeddes/ApTranscriptome).

@@

Figure 1. Matched expression type data for *A. carolinensis* (rows) and *A. picea* (columns). The color of each cell represent the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies. Blue cells are those with significant excesses, while red cells are those with significant absences.

Figure 2. Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures. Red bars and lines are for *A. carolinensis* while blue bars and lines are for *A. picea*.

Figure 3. Scatterplots of (A) constitutive expression in *A. picea* against degree of induction for upregulated transcripts in *A. carolinensis* and (B) constitutive expression in *A. carolinensis* against degree of induction for upregulated transcripts in *A. picea.*. For each species, the transcripts are split into expression category based on the ... The top row is for *High* transcripts, middle is for *Intermediate* transcripts and bottom row is for *Low* transcripts.

@@

>>>>

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis* across 12 temperature intervals that spanned their entire thermal breadth. We found that at least 8% of all genes were thermally responsive. The majority of upregulation was specific to low temperatures, and consistent with the *enhanced defense hypothesis* the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptation. To better understand the relationship between these two mechanisms, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. We found substantial differences in their patterns of gene expression. *Aphaenogaster picea* responded to extreme temperatures primarily through upregulation, with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes. These results suggest that adaptation to chronic high temperatures is not just a more extreme version of the acclimation response, but instead involves a qualitative shift in mechanisms from those used to cope with rarely encountered stressors.

@@

We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced defense hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced responses to provide greater protection from more frequently encountered stressors. This hypothesis would predict that the cool-adapted *A. picea* would induce more genes in response to low temperatures than would the warm-adapted *A. carolinensis*, which would induce more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should also be stronger for more commonly experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014]. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways. Finally, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, whereas transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation.

@@

pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
             big.mark = ",", split.tables = Inf, 
             caption = "Table of the number of thermally-responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatures (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species.")
```

@@

In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species. Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another. These shifts were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Figure S1) indicating that these differences are not simply due to neutral processes. The *enhanced defense* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species. The *enhanced defense hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response, whereas the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species. For each temperature extreme, we tested these predictions by focusing on the genes upregulated by the less-adapted species (*A. carolinensis* for low temperatures, *A. picea* for high temperatures) and testing whether those genes displayed the same response profile in the better-adapted species. As with the overall results, the patterns of association differed for the two temperature extremes. Genes upregulated at low temperatures in *A. carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from all other expression categories in *A. picea*, suggesting shared response pathways (Fig. 1A). In contrast, genes upregulated in response to high temperatures in *A. picea* (**High** and **Bimodal**) were only weakly associated with this same category in *A. carolinensis* (Fig. 1B). Instead, these genes were less likely than expected to be upregulated in any context, with a significant excess in the **Not Responsive** and **Intermediate** categories (Fig. 1B).

![Deviations from expected numbers of transcripts in matched observation data for transcript expression type data. The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit). Shown are **Low**, **High** and **Bimodal** expression types for (A) *A. picea* and (B) *A. carolinensis* against all expression types in the opposite species. The dashed boxes indicate the diagonals (e.g. **High** - **High**).](../results/matched_observations_Fig1.png)

The molecular processes in the set of transcripts that switched from upregulation in *A. picea* to downregulation in *A. carolinensis* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism. In addition, there were GO terms related to repair (e.g. histone modification), transcript regulation (e.g. positive regulation of cellular process) and cellular organization (e.g. transport, organelle organization). Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A. carolinensis*, primarily shuts down metabolism whereas the cooler-climate species, *A. picea*, increases metabolism and initiates damage repair responses.

@@

![Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures. Red bars and lines are for *A. carolinensis* while blue bars and lines are for *A. picea*.](../results/crit_temp_regulation_Fig2.png)

@@

We tested the *genetic assimilation hypothesis*, a shift from inducible to constitutive expression of stress-response genes, by comparing the log ratios of relative inducibility to relative constitutive expression. Specifically, this hypothesis predicted a negative relationship between the two. We found no evidence of such a relationship for either temperature extreme; transcripts upregulated at high temperatures in *A. picea* (Fig. 3A) or transcripts upregulated at low temperatures in *A. carolinensis* (Fig. 3B). In fact, for both comparisons we found a weakly positive relationship.

![Scatterplots of (A) log ratio of inducibility in *A. picea* over inducibility in *A. carolinensis* against log ration of constitutive expression in *A. carolinensis* over constitutive expression in *A. picea* for transcripts with **High** expression in *A. picea*, and (B) log ratio of inducibility in *A. carolinensis* over inducibility in *A. picea* against log ration of constitutive expression in *A. picea* over constitutive expression in *A. carolinensis*. Blue lines and confidence intervals are from are ordinary least squares regressions.](../results/genetic_assimilation_Fig3.png)

@@

The potential for many species to persist in face of climate change will depend in part upon their thermal tolerances. However, for most species little is known about how plasticity or adaptive changes in gene expression underlie thermal tolerance. By using a *reactionome* approach, we were able to quantitatively describe plasticity in transcript expression across a thermal gradient, and identify evolutionary changes in gene expression associated with shifts in thermal tolerance between the ant species *Aphaenogaster picea* and *A. carolinensis*. We found large-scale, non-linear patterns of gene expression changes in response to temperature, with both quantitative and qualitative differences between species, consistent with different mechanisms of thermal adaptation to low and high temperature extremes.

Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig. 2A). Moreover, the transcripts upregulated in *A. picea* included nearly the entire set upregulated in *A. carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* may induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the shorter growing season that would necessitate foraging under a broader range of temperatures.

In contrast to cold tolerance, the enhanced upper thermal limit in *A. picea* is best explained by the *tolerance* hypothesis. High temperatures were associated with significantly fewer upregulated transcripts in *A. carolinensis*, and a large proportion (361 of 1,238 transcripts) of those transcripts upregulated in *A. picea* were either downregulated or expressed at negligible levels overall in *A. carolinensis*. Similarly, significantly higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* towards the high thermal limit (Fig. 2B-D). This difference between species is also evident among candidate genes known to be involved in stress response. Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A. picea* whereas only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*. Altogether, these results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases, which may include constitutive defenses [@barshis2013],  enhanced proteome stability [@ghosh2010] or behavioral changes [@sunday2014] to tolerate thermal stress. These differences are in line with expectations that *A. carolinensis*, with a growing season over twice the length of its northern congener, may be better able to afford to restrain from foraging in suboptimal conditions. In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

The one hypothesis that did not receive support was the *genetic assimilation hypothesis*, which predicts that exposure to more frequent stressors will select for a shift from inducible to constitutive expression of stress-response transcripts. This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013], although in a short-term selection experiment for heat tolerance, Sikkink et al. [-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations. Both the genetic assimilation and tolerance routes routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair. Whether a particular taxon evolves one strategy over another may be related to the intensity, frequency and duration of temperature stress in a given environment.

Given the differences in the patterns of thermal responsiveness between species (Fig. 1, 2), it is worth noting a number of similarities. In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Table 1). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. Among closely-related *Drosophila* species, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. To the extent that the number of thermally responsive genes is related to additive genetic variation, our results also suggest that adaptability to high temperatures is more genetically limited than low temperatures, perhaps placing these northern species at risk of competitive replacement by more heat-tolerant southern species [@rodriguez-cabal2012]. The two species also showed similar numbers of genes with a **Bimodal** response (Table 1), indicating a generalized stress-response role. Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.  Only 15-20% of transcripts upregulated in response to temperature are bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig. 2). This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.

@@

Moreover, by characterizing responses across thousands of transcripts, the reactionome approach can help to distinguish selection from neutral drift in gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009]. While we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in , expression type categories or critical temperature thresholds as we do here (Fig. 1, Fig. S1). Thus, our method provides an example of how focusing on transcriptome-wide changes in gene expression, as opposed to identifying lists of differentially-expressed genes, can provide meaningful insight on the process of evolution.

@@

Illumina short-read sequence data is available from the NCBI Short Read Archive BioProject PRJNA260626. The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files are available at [TBD: Harvard Forest Archives]. Reproducible R script for analysis is available from GitHub [@stanton-geddes2014].

## Acknowledgements

Support made possible by NSF DEB Award #1136644 and the Vermont Genetics Network through Grant Number 8P20GM103449 from the INBRE Program of the National Institute of General Medical Sciences (NIGMS) of the National Institutes of Health (NIH). Its contents are solely the responsibility of the authors and do not necessarily represent the official views of NIGMS or NIH.

@@

Figure legends go here.

@@

Figures here.
